---
title: "Repeated-measures ANOVA"
author: ""
date: "`r Sys.Date()`"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [pomegranate_on_print.css]
nature: 
  highlightStyle: tomorrow-night-blue
  hilightLines: true
  highlightSpans: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.showtext=TRUE)
library(tidyverse) # load tidyverse package
library(MASS)
library(ggplot2)
library(ggthemes)
library(kableExtra)
library(flextable)
library(cowplot)
library(xaringanthemer)
#library(gg3D)
#library(leaflet)
library(maps)
library(fontawesome)
library(ggrepel)
library(DescTools)
library(renderthis)
library(ftExtra)
library(BayesFactor)
library(ggridges)
library(wesanderson)
library(lme4)

RM_1way_ex_df<-data.frame(Participant = factor(rep(c("P1", "P2", "P3", "P4"),
                                            each = 3)),
                          IV = factor(rep(c("A1", "A2", "A3"),
                                   4)),
                          DV = c(5, 6, 7,
                                 7, 13, 13,
                                 2, 4, 6,
                                 6, 9, 12))

DV<-RM_1way_ex_df$DV
IV<-RM_1way_ex_df$IV
Participant<- RM_1way_ex_df$Participant

#library(xaringan)
#decktape(
#  file = "Repeated_measures_ANOVA.html",
#  output = "Repeated_measures_ANOVA.pdf",
#  args = "--chrome-arg=--allow-file-access-from-files",
#  docker = FALSE,
# # version = "",
#  open = FALSE
#)
## To create pdf version:

# setwd("~/Documents/PSY 208/PSY207_208_24_25/Lectures")
# library(renderthis)

# renderthis::to_pdf("Repeated_measures_ANOVA.Rmd",  complex_slides = TRUE)

## To create PowerPoint version:

# renderthis::to_pptx("Repeated-measures_ANOVA.Rmd", complex_slides = TRUE)
```

### Repeated-measures Designs

.slightly-smaller[
Also known as **within-participants/subjects** or **Randomized Block**

> **block** frequently refers to **individuals in different conditions**, but could also refer to other **matching variables**

Each participant gets **each level** of a treatment

Advantages of **repeated-measures designs** (as first discussed in the context of repeated-measures $t$-tests):

> Increased power due to reducing the impact of individual differences

> Fewer observations required

> Reduced cost, resources
]

---
### Repeated-measures ANOVA: Assumptions



Assumptions of the repeated-measures design:

> **Normality**: the **residuals** are normally distributed 

>> For **repeated-measures**, residual calculation also accounts for **participant**/**block** differences


> **Additivity**: no **subject-by-treatment interaction**

>>Differences in conditions, on average, affect individuals in the same way.

---

### Repeated-measures ANOVA: Assumptions


> **Sphericity**: the **variances of the differences** between each **pair** of conditions are equal.

>> The **variances of the differences** are like ** $s_d$** from RM $t$-tests - just now there are *more of them*.


If any of these assumptions are violated, we may still proceed, it’s just a little more complicated.


---
### Repeated-measures ANOVA: Hypotheses

The null and alternative hypotheses for repeated-measures ANOVA **are the same as for the independent-groups ANOVA**:

$$H_0: \sigma^2_\alpha = 0$$

$$H_1: \sigma^2_\alpha>0$$

That is, **the null is that there is no variance associated with the treatment** $(\alpha)$ and **the alternative is that there is variance associated with the treatment** $(\alpha)$ on the population level.

The **difference** between repeated-measures and independent-groups ANOVA is **how we test the hypotheses**.

---

### Repeated-measures ANOVA Model

Assuming **additivity**:

** $$y_{ij}=\mu+\alpha_j+\pi_i+\epsilon_{ij}$$**

.center[
** $y_{ij}$**: observed value of the DV

** $\mu$**: population mean

** $\alpha_j$**: effect of the treatment $\alpha$ at each level $j$

** $\pi_i$**: effect of *individual* (or *block*) *differences*

** $\epsilon_{ij}$**: error (or, residuals)

]

---

### RM model *vs.* IG model

IG model:  $y_{ij}=\mu+\alpha_j$ ** $+\epsilon_{ij}$**

> population variance components: $\sigma^2_\alpha, \sigma^2_\epsilon$

RM model: $y_{ij}=\mu+\alpha_j$ ** $+\pi_i+\epsilon_{ij}$**

> population variance components: $\sigma^2_\alpha, \sigma^2_\epsilon, \sigma^2_\pi$

Important concept: the repeated-measures model **does not *add* variance** associated with individual differences - **it takes that variance *out*** of the error

> A smaller error term tends to mean a larger $F$ ratio – and therefore more power for rejecting $\sigma^2_\alpha=0$.

---

### RM ANOVA Calculation (Additive)

Almost all of the calculations for the Sums of Squares terms are the same for the repeated-measures ANOVA as they are for the independent-groups ANOVA, with **one additional step**.

>We are additionally going to calculate the **Sums of Squares associated with individual differences**: ** $SS_{participants}$**

> We will then take that value and subtract it from ** $SS_{within}$** to get ** $SS_{e}$** 

---

### RM One-way ANOVA Example

Four participants $(n=4)$, three levels of **Factor A**\*, $\alpha = 0.05$

```{r echo=FALSE}

RM_1way_ex_df %>% 
  pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  flextable() %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  align(align = "center", j = 2:4, part = "all") %>% 
  fontsize(size = 18, part = "header") %>%
  fontsize(size=16, part = "body") %>% 
  width(width = 1.75, unit = "in")
```

.footnote[
\* there is no **Factor B** here - just keeping our notation consistent between one-way and factorial ANOVA
]

---
### RM One-way ANOVA Example

$SS_{total}: \sum \left( y_{ij}-y_{\bullet \bullet}\right)^2$ - **This is the same as IG ANOVA**

```{r echo=FALSE}
GM = mean(RM_1way_ex_df$DV)

RM_1way_ex_df %>% 
    pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  mutate(A1resid = round(A1-GM, 2)) %>% 
  mutate(A2resid = round(A2-GM, 2)) %>% 
  mutate(A3resid = round(A3-GM, 2)) %>% 
  mutate(A1residsq = round(A1resid^2, 2)) %>% 
  mutate(A2residsq = round(A2resid^2, 2)) %>% 
  mutate(A3residsq = round(A3resid^2, 2)) %>% 
  relocate(c(A1resid, A1residsq), .after = A1) %>% 
  relocate(c(A2resid, A2residsq), .after = A2) %>% 
  relocate(c(A3resid, A3residsq), .after = A3) %>% 
  flextable() %>% 
  set_header_labels(Participant = "P",
                    A1resid = "$y_{i1}-y_{\\bullet \\bullet}$",
                    A1residsq = "$(y_{i1}-y_{\\bullet \\bullet})^2$",
                    A2resid = "$y_{i2}-y_{\\bullet \\bullet}$",
                    A2residsq = "$(y_{i2}-y_{\\bullet \\bullet})^2$",
                    A3resid = "$y_{i3}-y_{\\bullet \\bullet}$",
                    A3residsq = "$(y_{i3}-y_{\\bullet \\bullet})^2$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  bg(bg = "#648fff", j = c(4, 7, 10), part = "body") %>% 
  align(align = "center", j = 2:10, part = "all") %>% 
  fontsize(size = 18, part = "header") %>%
  fontsize(size=16, part = "body") %>% 
    add_header_row(values = c("",
                              "$y_{\\bullet\\bullet} = 7.5$"),
                 colwidths = c(1, 9)) %>% 
  align(align = "center", part = "header") %>% 
  set_table_properties(layout = "autofit") %>% 
  colformat_md(part = "header")
```

$SS_{total}= \sum \left( y_{ij}-y_{\bullet \bullet}\right)^2=139$

---
### RM One-way ANOVA Example

$SS_{A}: n\sum \left( y_{\bullet j}-y_{\bullet \bullet}\right)^2$ - **This is ALSO the same as IG ANOVA**

```{r echo=FALSE}
GM = mean(RM_1way_ex_df$DV)


RM_1way_ex_df %>% 
    pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  mutate(A1resid = round(mean(A1)-GM, 2)) %>% 
  mutate(A2resid = round(mean(A2)-GM, 2)) %>% 
  mutate(A3resid = round(mean(A3)-GM, 2)) %>% 
  mutate(A1residsq = round(A1resid^2, 2)) %>% 
  mutate(A2residsq = round(A2resid^2, 2)) %>% 
  mutate(A3residsq = round(A3resid^2, 2)) %>% 
  relocate(c(A1resid, A1residsq), .after = A1) %>% 
  relocate(c(A2resid, A2residsq), .after = A2) %>% 
  relocate(c(A3resid, A3residsq), .after = A3) %>% 
  flextable() %>% 
  set_header_labels(Participant = "P",
                    A1resid = "$y_{\\bullet 1}-y_{\\bullet \\bullet}$",
                    A1residsq = "$(y_{\\bullet 1}-y_{\\bullet \\bullet})^2$",
                    A2resid = "$y_{\\bullet 2}-y_{\\bullet \\bullet}$",
                    A2residsq = "$(y_{\\bullet 2}-y_{\\bullet \\bullet})^2$",
                    A3resid = "$y_{\\bullet 3}-y_{\\bullet \\bullet}$",
                    A3residsq = "$(y_{\\bullet 3}-y_{\\bullet \\bullet})^2$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  bg(bg = "#648fff", j = c(4, 7, 10), part = "body") %>% 
  align(align = "center", j = 2:10, part = "all") %>% 
  fontsize(size = 16, part = "header") %>%
  fontsize(size=12, part = "body") %>% 
      add_header_row(values = c("",
                              "$y_{\\bullet 1} = 5$",
                              "$y_{\\bullet 2} = 8$",
                              "$y_{\\bullet 3} = 9.5$"),
                 colwidths = c(1, 3, 3, 3)) %>% 
    add_header_row(values = c("",
                              "$y_{\\bullet\\bullet} = 7.5$"),
                 colwidths = c(1, 9)) %>% 
  align(align = "center", part = "header") %>% 
  set_table_properties(layout = "autofit") %>% 
  colformat_md(part = "header")
```

$SS_{A}= n\sum \left( y_{\bullet j}-y_{\bullet \bullet}\right)^2=42$

---
### RM One-way ANOVA Example

$SS_{within}: \sum \left( y_{i j}-y_{\bullet j}\right)^2$ - **This is the same as $SS_{within}$ or $SS_{e}$ for IG ANOVA**

```{r echo=FALSE}
GM = mean(RM_1way_ex_df$DV)


RM_1way_ex_df %>% 
    pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  mutate(A1resid = round(A1 - mean(A1), 2)) %>% 
  mutate(A2resid = round(A2 - mean(A2), 2)) %>% 
  mutate(A3resid = round(A3 - mean(A3), 2)) %>% 
  mutate(A1residsq = round(A1resid^2, 2)) %>% 
  mutate(A2residsq = round(A2resid^2, 2)) %>% 
  mutate(A3residsq = round(A3resid^2, 2)) %>% 
  relocate(c(A1resid, A1residsq), .after = A1) %>% 
  relocate(c(A2resid, A2residsq), .after = A2) %>% 
  relocate(c(A3resid, A3residsq), .after = A3) %>% 
  flextable() %>% 
  set_header_labels(Participant = "P",
                    A1resid = "$y_{i1}-y_{\\bullet 1}$",
                    A1residsq = "$(y_{i1}-y_{\\bullet 1})^2$",
                    A2resid = "$y_{i2}-y_{\\bullet 2}$",
                    A2residsq = "$(y_{i2}-y_{\\bullet 2})^2$",
                    A3resid = "$y_{i3}-y_{\\bullet 3}$",
                    A3residsq = "$(y_{i3}-y_{\\bullet 3})^2$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  bg(bg = "#648fff", j = c(4, 7, 10), part = "body") %>% 
  align(align = "center", j = 2:10, part = "all") %>% 
  fontsize(size = 16, part = "header") %>%
  fontsize(size=16, part = "body") %>% 
      add_header_row(values = c("",
                              "$y_{\\bullet 1} = 5$",
                              "$y_{\\bullet 2} = 8$",
                              "$y_{\\bullet 3} = 9.5$"),
                 colwidths = c(1, 3, 3, 3)) %>% 
  align(align = "center", part = "header") %>% 
  set_table_properties(layout = "autofit") %>% 
  colformat_md(part = "header")
```

$SS_{within}= \sum \left( y_{i j}-y_{\bullet j}\right)^2=97$

---
### RM One-way ANOVA Example

$SS_{participants}: \sum j\left( y_{i \bullet}-y_{\bullet \bullet}\right)^2$ - **THIS IS NEW**

First, we will calculate **participant means $(y_{i\bullet})$** across the $j$ conditions

```{r echo=FALSE}
GM = mean(RM_1way_ex_df$DV)


RM_1way_ex_df %>% 
    pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  group_by(Participant) %>% 
  mutate(Pmean = mean(c(A1, A2, A3))) %>% 
  flextable() %>% 
  set_header_labels(Participant = "P",
                    Pmean = "$y_{i \\bullet}$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  bg(bg = "#fe6100", j = c(5), part = "body") %>% 
  align(align = "center", j = 2:5, part = "all") %>% 
  fontsize(size = 24, part = "header") %>%
  fontsize(size=24, part = "body") %>% 
  align(align = "center", j = 2:5, part = "header") %>% 
  width(width = 1.5, unit = "in") %>% 
  colformat_md(part = "header")
```

---
### RM One-way ANOVA Example

$SS_{participants}: \sum j\left( y_{i \bullet}-y_{\bullet \bullet}\right)^2$ - **THIS IS NEW**

Then, we will add the squared difference between the **participant means $(y_{i\bullet})$** and **the grand mean $(y_{\bullet \bullet})$** or each of the $j$ conditions.

```{r echo=FALSE}
GM = mean(RM_1way_ex_df$DV)


RM_1way_ex_df %>% 
    pivot_wider(names_from = "IV",
              values_from = "DV") %>% 
  group_by(Participant) %>% 
  mutate(Pmean = mean(c(A1, A2, A3))) %>% 
    mutate(A1resid = round(Pmean - GM, 2)) %>% 
  mutate(A2resid = round(Pmean - GM, 2)) %>% 
  mutate(A3resid = round(Pmean - GM, 2)) %>% 
  mutate(A1residsq = round(A1resid^2, 2)) %>% 
  mutate(A2residsq = round(A2resid^2, 2)) %>% 
  mutate(A3residsq = round(A3resid^2, 2)) %>% 
  relocate(c(A1resid, A1residsq), .after = A1) %>% 
  relocate(c(A2resid, A2residsq), .after = A2) %>% 
  relocate(c(A3resid, A3residsq), .after = A3) %>% 
  flextable() %>% 
  set_header_labels(Participant = "P",
                    Pmean = "$y_{i \\bullet}$",
                    A1resid = "$y_{1 \\bullet}-y_{\\bullet \\bullet}$",
                    A1residsq = "$(y_{1 \\bullet}-y_{\\bullet \\bullet})^2$",
                    A2resid = "$y_{2 \\bullet}-y_{\\bullet \\bullet}$",
                    A2residsq = "$(y_{2 \\bullet}-y_{\\bullet \\bullet})^2$",
                    A3resid = "$y_{3 \\bullet}-y_{\\bullet \\bullet}$",
                    A3residsq = "$(y_{3 \\bullet}-y_{\\bullet \\bullet})^2$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
  bg(bg = "#fe6100", j = 10, part = "body") %>%
  bg(bg = "#648fff", j = c(4, 7, 10), part = "body") %>% 
  align(align = "center", j = 2:11, part = "all") %>% 
  fontsize(size = 16, part = "header") %>%
  fontsize(size=16, part = "body") %>% 
    add_header_row(values = c("",
                              "$y_{\\bullet\\bullet} = 7.5$"),
                 colwidths = c(1, 10)) %>% 
  align(align = "center", j = 2:11, part = "header") %>% 
  set_table_properties(layout = "autofit") %>% 
  colformat_md(part = "header")
```

$$SS_{participants}: \sum j\left( y_{i \bullet}-y_{\bullet \bullet}\right)^2=87$$
---

### RM One-way ANOVA Example

The sums of squares associated with participants are **literally subtracted** from what *was* the error for the IG ANOVA:

$$SS_{residual}=SS_{within}-SS_{participants}=97-87=10$$
---

### RM One-way ANOVA Table

```{r echo=FALSE}
Source<-c("Factor A",
          "Participants",
          "Error\n(Residuals)",
          "Total")

df <- c("$j-1$",
        "$n-1$",
        "$(j-1)(n-1)$",
        "$jn-1$")

SS <-c("$SS_A$",
       "$SS_p$",
       "$SS_{resid}$",
       "$SS_{total}$")

MS<-c("$MS_A$",
       "$MS_p$",
      "$MS_{resid}$",
       NA)

Fratio<-c("$MS_A/MS_e$",
          "Please ignore me",
          NA,
          NA)

EMS<-c("$n\\sigma^2_\\alpha + \\sigma^2_\\epsilon$",
       "$j\\sigma^2_\\pi + \\sigma^2_\\epsilon$",
       "$\\sigma^2_{\\epsilon}$",
       NA)


ANOVA_table<-data.frame(Source,
                        df,
                        SS,
                        MS,
                        Fratio,
                        EMS)

ANOVA_table %>% 
  flextable() %>% 
  set_header_labels(df = "$df$",
                    Fratio = "$F$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
#  bg(bg = "#648fff", j = c(3, 6, 9), part = "body") %>% 
  align(align = "center", j = 2:6, part = "all") %>% 
  fontsize(size = 18, part = "header") %>%
  fontsize(size=16, part = "body") %>% 
  width(width = 1.5, unit = "in") %>% 
  colformat_md(part = "all") 
```

.slightly-smaller[
Note: we *could* calculate an $F$-ratio for participants: it would be $MS_{participants}/MS_e$, and software packages (including `R`) will calculate it for you. However, we don't care about it because *we already know* that individual differences exist.
]
---
### RM One-way ANOVA Table

Here's the simplest way to do it in `R`. Don't get **too attached** to that $F$-ratio:

```{r}
summary(aov(DV~IV+Participant, 
            data = RM_1way_ex_df))
```

---

### RM One-way ANOVA Table

Here's another method that introduces the concept of **hierarchical regression modeling**

.pull-left[
.slightly-smaller[

At right is a *scatterplot* of the observed data: 

> the $x$-axis gives the **levels of factor A** 

> the $y$-axis gives the **observed values of the DV**

> the **color** of the dots varies by **participant**

]
]
.pull-right[

```{r echo=FALSE,  fig.align='center'}
RM_1way_ex_df %>% 
  ggplot(aes(x=IV, y = DV, color = Participant))+
  geom_point(size=10)+
  scale_color_manual(values = wes_palette(name = "Royal1",
                                          n = 4,
                                          type = "discrete"))+
  theme_tufte(ticks=FALSE)+
  theme_xaringan(css_file = "pomegranate_on_print.css")+
  theme(legend.position = "bottom")

```

]
---

### RM One-way ANOVA Table

.pull-left[


> This is a visualization of the concept of RM ANOVA as a **regression model** where each participant has a **different intercept** 

> The **overall slope** is the same! That's the **levels of Factor A** affecting the DV **consistently**



]

```{r echo = FALSE}

RM_MLM <- lmer(DV ~ IV + (1|Participant), data = RM_1way_ex_df)

A_means <- RM_1way_ex_df %>% 
  group_by(IV) %>% 
  summarise(A_means = mean(DV)) %>% 
  dplyr::select(A_means) %>% 
  pull()

P_means <- RM_1way_ex_df %>% 
  group_by(Participant) %>% 
  summarise(P_means = mean(DV)) %>% 
  dplyr::select(P_means) %>% 
  pull()

intercept = fixef(RM_MLM)[1]

slope = (max(A_means) - min(A_means))/3 

p_intercepts = unname(unlist(ranef(RM_MLM)))

```

.pull-right[

```{r echo=FALSE,  fig.align='center'}

colors = wes_palette(name = "Royal1",
                     n = 4,
                     type = "discrete")
RM_1way_ex_df %>% 
  ggplot(aes(x=IV, y = DV, color = Participant))+
  geom_point(size=10)+
  scale_color_manual(values = colors)+
  theme_tufte(ticks=FALSE)+
  theme_xaringan(css_file = "pomegranate_on_print.css")+
  theme(legend.position = "bottom")+
  geom_abline(slope = slope, intercept = intercept + p_intercepts[1],
              color = colors[1]) +
  geom_abline(slope = slope, intercept = intercept + p_intercepts[2],
              color = colors[2]) +
geom_abline(slope = slope, intercept = intercept + p_intercepts[3],
              color = colors[3]) +
geom_abline(slope = slope, intercept = intercept + p_intercepts[4],
              color = colors[4])
  

```

]

---

### RM One-way ANOVA Table

We can get the **same $F$-ratio** using a **multi-level regression model**\*

.slightly-smaller[

> The effect of **Factor A** (level 1) is nested within the effect of **participant differences** (level 2)

> The term `(1|Participant)` means that each participant represents a **random intercept** in a multi-level regression analysis

]

.smaller-output[

```{r}
library(lme4) #for multilevel modeling
anova(lmer(DV ~ IV + (1|Participant), data = RM_1way_ex_df))
```

]

***
.footnote[

.slightly-smaller[

\*this method will be quite helpful for more complex designs
]
]


---
### Effect Size

The value of $\eta^2$ is calculated and interpreted in the exact same way for the one-way RM ANOVA as it was for the one-way IG ANOVA:

$$\eta^2=\frac{SS_A}{SS_{total}}$$
$\omega^2$ is also very similar: it is still the ratio of the population variance component associated with Factor A divided by the sum of the other population variance components. 

> The difference is that there is **another** population variance component in the RM model: ** $\sigma^2_{\pi}$**:

$$\omega^2=\frac{\sigma^2_\alpha}{\sigma^2_\alpha+\sigma^2_\pi+\sigma^2_\epsilon}$$
---

### Effect size

The guide to estimating $\sigma^2_\pi$ comes from the EMS for **participants**:

$$\text{EMS}_{participants} = j\sigma^2_\pi + \sigma^2_\epsilon$$
Our estimate of $\sigma^2_\pi$ is therefore:

$$\widehat{\sigma^2_\pi}=\frac{\text{MS}_{participants}-\text{MS}_{residuals}}{j}$$
Note: $\widehat{\sigma^2_\pi}$ is **never corrected** because *participants* are **always** considered **random**!

---

### About those RM Assumptions

> **Additivity and Non-additivity**

The **additivity assumption** implies that each participant responds to each level of the IV factor(s) in an approximately equal way.

If we have reason to believe that there *is* a **participant-by-treatment interaction** - *and we are willing to deal with the consequences of that* - there is a **nonadditive model** for repeated-measures designs.

The one-way **nonadditive model** is:

$$y_{ij}=\mu+\alpha_j+\pi_i+\epsilon_{ij}+\alpha\pi_{ij}$$
where $\alpha_{\pi_{ij}}$ represents the **participant-by-treatment** (or, **blocks-by-treatment**) **interaction**.

---

### RM One-way ANOVA Table (nonadditive)

```{r echo=FALSE}
Source<-c("Factor A",
          "Participants",
          "A*Participants",
          "Total")

df <- c("$j-1$",
        "$n-1$",
        "$(j-1)(n-1)$",
        "$jn-1$")

SS <-c("$SS_A$",
       "$SS_p$",
       "$SS_{e}$",
       "$SS_{total}$")

MS<-c("$MS_A$",
       "$MS_p$",
      "$MS_{e}$",
       NA)

Fratio<-c("$MS_A/MS_{e}$",
          NA,
          NA,
          NA)

EMS<-c("$n\\sigma^2_\\alpha + \\sigma^2_{\\alpha\\pi}+\\sigma^2_\\epsilon$",
       "$j\\sigma^2_\\pi + (1-j/J)\\sigma^2_{\\alpha\\pi}+\\sigma^2_\\epsilon$",
       "$\\sigma^2_{\\epsilon}$",
       NA)


ANOVA_table<-data.frame(Source,
                        df,
                        SS,
                        MS,
                        Fratio,
                        EMS)

ANOVA_table %>% 
  flextable() %>% 
  set_header_labels(df = "$df$",
                    Fratio = "$F$") %>% 
  color(color = "white", part = "all") %>% 
  bg(bg = "#fe6100", part = "header") %>% 
#  bg(bg = "#648fff", j = c(3, 6, 9), part = "body") %>% 
  align(align = "center", j = 2:6, part = "all") %>% 
  fontsize(size = 14, part = "header") %>%
  fontsize(size=14, part = "body") %>% 
  set_table_properties(layout = "autofit") %>% 
  colformat_md(part = "all") 
```

.slightly-smaller[
There's not an appreciable difference for the **one-way** design between the **additive** and **nonadditive** analysis except for the **interpretation** of the nature of the error term.]

The difference will come in **factorial repeated-measures designs** 
---

### RM One-way ANOVA Table (nonadditive)

```{r}
summary(aov(DV~IV+ Participant + Error(IV:Participant), 
            data = RM_1way_ex_df))
```

---

### Issues with Nonadditive Analyses

.slightly-smaller[

The **additivity assumption** can be statistically assessed with **Tukey's test of additivity**


```{r}
asbio::tukey.add.test(DV, IV, Participant)
```


Nonadditive analyses:

> Have generally reduced power (Kirk, 1995)

> Are weird to explain

> Make it difficult to calculate effect sizes

So, best to **skip it if you can**.
]
---

### Sphericity

Unlike the other assumptions of parametric tests covered thus far, **violations of Sphericity** ***increase the likelihood of type-I errors***

The proper $F$-distribution for testing $H_0$ is called the **Adjusted $F$**, and it has **reduced degrees of freedom in the numerator and denominator** relative to the regular $F$.

> The extent to which the $df$ are reduced depends on the **severity of the violations** and is called ** $\epsilon$** (statisticians can't name things well) and we're going to let the software take care of calculating it for us.

> $\epsilon$ ranges from $\frac{1}{j-1}$ (maximum sphericity violation) to $1$ (no sphericity violation)

---

### Sphericity

To get the adjusted $F$, we multiply the usual $df_{numerator}$ and $df_{denominator}$ by $\epsilon$.

> That means that the $[df_{num}, df_{denom}]$ for the adjusted $F$ are somewhere between: $[1, n-1]$ and $[j-1, (j-1)(n-1)]$

> Note: If we assume the **minimum possible value of $\epsilon$**\*, then the $F$-distribution has $[df_{num}, df_{denom}]= [1, n-1]$ and the test is called the **conservative $F$-test**.


.footnote[

the *minimum* $\epsilon$ means sphericity is *maximally* violated

]


---

### Sphericity

The **Geisser-Greenhouse Adjusted $F$-test** is probably the most common **Adjusted- $F$ procedure** for repeated-measures designs in which sphericity and/or compound symmetry assumptions may be violated. It's a little **too conservative** when $\epsilon$ is relatively large.

The **Huynh-Feldt** adjustment tweaks the Geisser-Greenhouse test to be a little less conservative (and therefore more powerful), but doesn't do as good a job when departures from sphericity are severe.

```{r eval = FALSE}
library(ez)
ezANOVA(data=RM_1way_ex_df,
        dv = .(DV),
        wid = .(Participant),
        within = .(IV),
        type = 3)
```

---

### Sphericity

Because the **adjustment** of $F$ is **directly related to the extent of the violation** of sphericity, a good approach is to **just use an adjusted $F$ in the first place**.


```{r echo = FALSE}
library(ez)
ezANOVA(data=RM_1way_ex_df,
        dv = .(DV),
        wid = .(Participant),
        within = .(IV),
        type = 3)
```


> The general recommendation is: use GG when `GGe` $\le 0.75$, HF otherwise.
---

### Nonparametric RM ANOVA

The nonparametric repeated-measures one-way ANOVA is known as the **Friedman Test**.

```{r}
friedman.test(DV,
              IV,
              Participant)
```

For this example, $p < 0.05$ indicates that **there is a signficant effect of the IV**.
---
### Bayesian RM ANOVA

The repeated-measures ANOVA using the `anovaBF()` function is nearly identical to the independent-groups ANOVA


> We include `Participant` in the model

> We indicate that `Participant` is the identifier for the blocking variable using `whichRandom`




```{r}
RM_Bayes_AOV <- anovaBF(DV~IV + Participant, 
                        data = RM_1way_ex_df,
                        whichRandom = "Participant",
                        progress = FALSE)
```

---

### Bayesian RM ANOVA

For the example data, the Bayes Factor for a model that includes **the IV and Participant** as predictors vs. a (null model) that includes **only Participants** ...

```{r}
RM_Bayes_AOV
```

... is $\approx 6$, indicating **substantial strength of evidence**.
